OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA

### 2018-11-14 phil_paper/jmp_qje.tex
better intro

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a 
Improving access to piped water generates substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  However, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, government subsidies are often limited by high costs of public funds and risk of regulatory capture in these countries (\cite{laffont2005regulation}).  Instead, prices must often be set to fully recover costs.  A popular solution is to subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities also increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in South Africa and C{\^o}te d'Ivoire predict gains in both access and total welfare from increasing marginal prices (\cite{szabo2015value,diakite2009proposal}).

However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.  Land-tenure issues, credit-constraints, and administrative costs often prevent disadvantaged households from purchasing their own connections (\cite{devoto2012happiness}).  Also, many cities maintain high upfront fees despite policy recommendations otherwise.  As a result, 70\% of African utilities report frequent water sharing between households (\cite{keener2010provision}) ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of informal sharing, previous research has not yet examined the importance of informal sharing for optimal pricing policy.

This paper analyzes the welfare effects of pricing policies in the presence of informal water sharing.  I build a model of household water demand that includes both (1) the upfront choice of whether to purchase a connection, share with a neighbor, or use non-piped sources as well as (2) monthly consumption choices.  I also collect new survey evidence on informal sharing contracts to inform several features of the model design.  The structural estimation leverages administrative data from 1.5 million water connections in Manila, which link to survey data on households owning and sharing connections.

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover both fixed and marginal costs while maximizing consumer surplus.  I also allow the regulator to place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have lower marginal prices, but varying quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.  The model collapses all non-piped sources into a single ``vendor'' option.}  The opposite approach --- high fixed fees and low marginal prices --- improves both access and welfare even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements allow small-users to access piped water from their neighbors at low marginal prices.  Shared connections also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.

%By using from their neighbors, smaller-users also benefit from low marginal prices and avoid sourcing from non-piped water.  
% Non-linear pricing helps policymakers achieve both efficiency (by targeting more inelastic households with higher prices) as well as equity (by targeting poorer households with lower prices).
% excluding many smaller-users from piped water.\footnote{In the model, households can substitute to a vendor option that encompasses all non-piped sources of water, which commonly include deepwells, tanker trucks, and water refilling stations.}  
%The model is designed to match several features from new survey evidence on sharing contracts that I fielded as part of this study.

%In the context of Manila, I find that the most disadvantaged households are also most likely to fetch water from their neighbors.
%In Manila, I establish that 23\% of households access water through a neighbor's connection by connecting plumbing directly to the meter (\input{tables/access_pipes}\unskip\%), extending plastic pipes (\input{tables/access_single_tap}\unskip\%), or fetching water with containers (\input{tables/access_fetch}\unskip\%).

Manila

To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2010 to 2015 with detailed survey data on the number of households and people using each water connection.

With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.

I find that shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.

%At the same time, informal sharing networks serve as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.

%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage .

Welfare impacts depend not only on heterogeneity in demand for water across the population, but also on costs associated with forming and maintaining water sharing relationships.  [DONT LIKE THIS SENTENCE ANYMORE]

To evaluate the impacts of different pricing policies, I build a model of household water demand that includes both initial source choices as well as monthly consumption choices.

This model expands on the demand estimation literature for public utilities that focuses almost exclusively on quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).  I specify a sequential game where households choose whether to connect individually, share with their neighbors, or purchase from a water vendor.  For sharing households, I develop a simple model of the water sharing contract that captures two key features of these relationships: (1) households face challenges in observing each other's demand and (2) households often split the monthly water bill evenly.  The model characterizes monthly consumption choices when households face non-linear budget constraints and imperfect control over their exact consumption levels.  This approach extends previous models of optimization over non-linear budget constraints by allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

The estimation combines water source and consumption decisions to structurally recover preferences and unobserved costs of water sharing.  To include reduced-form approaches to identification within the context of a structural model, I divide the estimation into three steps.  The first step uses variation in the non-linear price schedule (both across and within households) to identify water demand for each household using monthly consumption data.  To measure the shadow costs of fetching water from a shared connection, the second step analyzes a quasi-experiment where households face leaks in their service pipes, forcing them to substitute to a neighbor's connection.  The model maps the extent to which these households are able to offset their consumption to a neighbor's connection into an estimate of the marginal costs of fetching water.  This exercise recovers sizable sharing costs on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.  This quasi-experimental approach builds on a descriptive literature pointing to potential costs of sharing water (\cite{whittington1992possible,nauges2006water}).  On top of an initial connection fee and monthly service fee, households may face unobserved fixed costs in owning a connection, which may include application costs, plumbing maintenance, as well as land-tenure issues in qualifying for a water connection.  The third step estimates these fixed costs by simulating the sequential game to match cross-sectional variation in water source choices.  This strategy provides a structural, revealed-preference approach to recovering unobserved fixed costs, complementing previous reduced-form evidence in \cite{devoto2012happiness}.

Combining the model and estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,wsup,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensuring that the water provider exactly covers costs.  This policy reduces access by incentivizing households to substitute toward water vendors, which mainly include deepwells and water refilling stations.

@@

Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus or \input{tables/tpt_improvement_income_sharing_group}\unskip\% of household income.  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections to install and maintain.  
%Yet, since sharing involves costly redistribution of water, the optimal tariff subsidizes individual connections on the margin by pricing above marginal cost and setting the fixed fee below the cost of installing and maintaining each connection.

Finally, I compare the optimal tariff with a social tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, social tariffs are ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, social tariffs achieve substantial redistribution, over doubling consumer surplus for small users.  This evidence suggests that the presence of informal sharing networks constrains non-linear pricing as an effective tool for redistribution.

@@

@@

This section describes an extension of the model to allow for exogenous interruptions in a water source that induce households to substitute toward another water source.  Exogenous interruptions produce useful variation in the types of water sources used by each household, serving an important role in estimating the hassle costs of transporting water from a neighbor.  Due to shifts in the geology, road construction, or aging infrastructure, households sometimes experience underground breaks in the service pipe between the main line and the water meter.  These leaks result in repair costs and water bills borne by households that can be large enough to force affected households to switch to an alternative water source.

@@

This paper demonstrates that pricing policies intended to increase water access especially among the poor can have unintended consequences when low-income households can simply access the piped water network through their neighbors.  The counterfactual exercises find that a simple two-part tariff structure outperforms the increasing block tariff currently implemented in Manila in terms of overall access to water, total welfare, and even welfare among the bottom half of water users.  In this counterfactual, high fixed costs of accessing an individual connection allow for sharing networks to play an important role in extending water access despite sizable hassle costs.  I also find that complex non-linear price schedules are largely ineffective in terms of achieving both efficiency and equity goals when sharing networks are readily available.

>>>>

\section{Introduction}
 % \footnote{\cite{laffont2005regulation} emphasizes high costs of public funds in developing countries (with a deadweight loss of 2.5 for the Philippines) due to limited ability to enter long-term contracts with government, credit constraints, inefficient taxation, and corruption.}
 % To ensure widespread access while covering costs, a 
Improving access to piped water can generate substantial health benefits\footnote{\cite{zhang2016long,kosec2014child,gamper2010impact}, and \cite{galiani2005water} focus on health improvements in developing countries while \cite{cutler2005role} look historically at the US.} as well as improvements in household well-being through time savings and reduced intra-household conflict.\footnote{\cite{devoto2012happiness} provide water through a randomized-controlled trial in Morocco finding few health effects but large benefits to household welfare.}  However, the share of urban households without piped water has remained stable at 5\% for middle income countries and at 20\% for low income countries (World Bank [2015]).  While external health and other benefits may merit subsidizing piped water, government subsidies are often limited by high costs of public funds and risk of regulatory capture in these countries (\cite{laffont2005regulation}).  In many cases, prices must be set to fully recover costs.  A popular solution is to subsidize upfront connection fees with high marginal prices for monthly usage.\footnote{\cite{mcintosh2003asian} and \cite{komives2006distributional} from the World Bank and Asian Development Bank argue for low connection fees.  Using a randomized controlled trial from Morocco, \cite{devoto2012happiness} confirm that modest assistance with connection fees can lead to increases in new water connections.}  Over 70\% of developing cities also increase marginal prices with monthly usage so that large users cross-subsidize access and usage for smaller, poorer users.\footnote{This strategy is also commonly referred to as ``Increasing Block Tariffs'' or  ``Convex Pricing'' (\cite{borenstein2012redistributional}). \cite{hoque2013state} and \cite{boland2000political} provide overviews of the use of increasing marginal prices in the water sector in developing countries.  The statistic comes from author calculations with data kindly provided by \cite{hoque2013state}.}  These pricing approaches have been shown to work well when each water connection serves a single household.  Recent empirical papers in South Africa and C{\^o}te d'Ivoire predict gains in both access and total welfare from increasing marginal prices (\cite{szabo2015value,diakite2009proposal}).  However, many households in developing cities share connections with their neighbors, which can have large implications for water pricing.

Land-tenure issues, credit-constraints, and administrative costs often prevent disadvantaged households from purchasing their own connections (\cite{devoto2012happiness}).  Also, many cities maintain high upfront fees despite policy recommendations otherwise (\cite{hoque2013state}).  As a result, 70\% of African utilities report frequent water sharing between households (\cite{keener2010provision}) ranging from fetching from neighbors with hoses and buckets in Tangiers, Morocco (\cite{devoto2012happiness}) to drawing from single taps that each serve compounds in Kumasi, Ghana (\cite{whittington1992possible}).\footnote{\cite{nauges2006water} finds further evidence of widespread water sharing in Sri Lanka.}  While documenting the prevalence of informal sharing, previous research has not yet examined the importance of informal sharing for optimal pricing policy.

This paper analyzes the welfare effects of pricing policies in the presence of informal water sharing.  I build a model of household water demand that includes both (1) the upfront choice of whether to purchase a connection, share with a neighbor, or use non-piped sources as well as (2) monthly consumption choices.  I also collect new survey evidence on informal sharing contracts to inform several features of the model design.  The structural estimation leverages administrative data from 1.5 million water connections in Manila from 2010 to 2015, which link to additional survey data on households sharing connections.

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have lower marginal prices, but varying quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.  The model collapses all non-piped sources into a single ``vendor'' option.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access and welfare even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements allow small-users to access piped water from their neighbors for even lower marginal prices than those offered by non-piped sources.  Shared connections also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{Cost savings are passed on to consumers through lower prices.}

The welfare impacts of these pricing policies depend not only on household-level demand, but also on the relative costs of sharing water; these may include both marginal costs (ie. time spent fetching from neighbors or reduced pressure from re-piping water) as well as savings from splitting the fixed costs of a connection.  Measuring demand and sharing costs presents two central challenges: how can we recover household-level demand when we only observe water consumption for each connection, and how can we separately identify the costs of sharing from preferences for water?  The empirical approach addresses these challenges in three steps.

The first step builds a tractable model of water sharing to estimate household-level demand from monthly usage data for each connection.  Informed by a new survey, this model embeds how (1) households face challenges in observing each other's monthly demand and (2) households often split the monthly water bill evenly.  These assumptions on sharing behavior ensure that connection-level demand can be expressed as a sum of underlying household demands even in the presence of non-linear prices.  I use variation in the non-linear price schedule (both across and within connections) to identify water demand for each connection before disaggregating preferences to the household-level.  This approach extends previous models of optimization over non-linear prices by including sharing and allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

To estimate the marginal costs of sharing water, the second step analyzes a quasi-experiment where breaks in service pipes force some households to switch from their own water connections to neighbors' connections.  I then measure how much less water households use from their neighbors' connections and attribute this gap to the marginal costs of sharing.  This exercise recovers sizable marginal costs on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.\footnote{This revealed-preference approach builds on a descriptive literature that identifies potential sources of water sharing frictions (\cite{whittington1992possible,nauges2006water}).}

The third step maps the share of households using from each source --- connecting alone, sharing with a neighbor, or using from a non-piped vendor --- into an estimate of each source's relative fixed cost.  The empirical approach leverages the assumption that given water preferences (from step one) and marginal sharing costs (from step two), any remaining variation in water source choices must be due to differences in their fixed costs.  Since sharing decisions are strategic in nature, I specify a simple game where households sequentially choose their sources and agree to split any surplus from sharing.  This revealed-preference approach to estimating fixed costs is especially relevant in a developing context because on top of the initial connection and monthly service fees, households often face fixed costs that are difficult to observe, including permitting fees, plumbing maintenance, as well as land-tenure issues in qualifying for a water connection.  In fact, I estimate that the connection and service fees account for only around half of the total fixed costs facing households.  This structural approach complements previous experimental evidence of fixed costs presented in \cite{devoto2012happiness}.  By considering both source and quantity choices, this model also expands on the demand estimation literature for public utilities that has so far focused almost exclusively on monthly quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).

%The third step estimates these fixed costs by simulating the sequential game to match cross-sectional variation in water source choices.

With these estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensuring that the water provider exactly covers costs.  This policy actually reduces access by incentivizing households to substitute toward non-piped vendors.  Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus.\footnote{With quasi-linear utility, consumer surplus for a given tariff is calculated as the difference between household utility under this tariff and household utility from using the non-piped, vendor option.}  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections to install and maintain.  Finally, I compare the optimal tariff with a social tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, social tariffs are ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, social tariffs achieve substantial redistribution.  These findings suggest that informal sharing networks constrain the extent to which non-linear prices can redistribute surplus.

%Yet, since sharing involves costly redistribution of water, the optimal tariff subsidizes individual connections on the margin by pricing above marginal cost and setting the fixed fee below the cost of installing and maintaining each connection.

% analyzes a quasi-experiment where households face leaks in their service pipes, forcing them to substitute to a neighbor's connection.

% The model maps the extent to which these households are able to offset their consumption to a neighbor's connection into an estimate of the marginal costs of fetching water.  This exercise recovers sizable sharing costs on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.  This quasi-experimental approach builds on a descriptive literature pointing to potential costs of sharing water (\cite{whittington1992possible,nauges2006water}).

% In Manila, I establish that 23\% of households access water through a neighbor's connection by connecting plumbing directly to the meter (\input{tables/access_pipes}\unskip\%), extending plastic pipes (\input{tables/access_single_tap}\unskip\%), or fetching water with containers (\input{tables/access_fetch}\unskip\%).  Time spent fetching water or reduced pressure-levels from informally re-piping both increase the marginal cost of sharing.  
%--- what is the relative fixed cost of

%combines monthly usage to recover household-level demand.
% recover household-level from connection usage,

% I develop a simple model of the water sharing contract that captures two key features of these relationships informed by survey data:

% The model characterizes monthly consumption choices when households face non-linear budget constraints and imperfect control over their exact consumption levels.

% Greater demand heterogeneity increases the scope for usage-based price discrimination to both redistribute surplus and efficiently raise revenue.

% Sharing costs determine how households can respond by changing their water source alongside their consumption patterns.

%  while low sharing costs allow households

% To evaluate the impacts of different pricing policies, I build a model of household water demand that includes both initial source choices as well as monthly consumption choices.

%For sharing households, I develop a simple model of the water sharing contract that captures two key features of these relationships: (1) households face challenges in observing each other's demand and (2) households often split the monthly water bill evenly.  The model characterizes monthly consumption choices when households face non-linear budget constraints and imperfect control over their exact consumption levels.  This approach extends previous models of optimization over non-linear budget constraints by allowing households to anticipate unobserved shocks to their demand (\cite{moffitt1986econometrics,burtless1978effect}).

% The estimation combines water source and consumption decisions to structurally recover preferences and unobserved costs of water sharing.  To include reduced-form approaches to identification within the context of a structural model, I divide the estimation into three steps.  The first step uses variation in the non-linear price schedule (both across and within households) to identify water demand for each household using monthly consumption data.

@@

@@

%By using from their neighbors, smaller-users also benefit from low marginal prices and avoid sourcing from non-piped water.  
% Non-linear pricing helps policymakers achieve both efficiency (by targeting more inelastic households with higher prices) as well as equity (by targeting poorer households with lower prices).
% excluding many smaller-users from piped water.\footnote{In the model, households can substitute to a vendor option that encompasses all non-piped sources of water, which commonly include deepwells, tanker trucks, and water refilling stations.}  
%The model is designed to match several features from new survey evidence on sharing contracts that I fielded as part of this study.
%In the context of Manila, I find that the most disadvantaged households are also most likely to fetch water from their neighbors.
%In Manila, I establish that 23\% of households access water through a neighbor's connection by connecting plumbing directly to the meter (\input{tables/access_pipes}\unskip\%), extending plastic pipes (\input{tables/access_single_tap}\unskip\%), or fetching water with containers (\input{tables/access_fetch}\unskip\%). 
% Manila 
% To measure these features, I merge administrative data covering 1.5 million water connections in Manila from 2010 to 2015 with detailed survey data on the number of households and people using each water connection.  
% With comprehensive geographic coverage of half of Manila and a variety of price variation, these data allow for a rich characterization of water demand.  Additional variation in water source choices identify the relative fixed and marginal costs associated with sharing water with a neighbor.  
% I find that shared connections allow the city to save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households at low marginal prices.  
%At the same time, informal sharing networks serve as resale markets for water, undermining the extent to which cities can redistribute surplus through increasing marginal prices.
%  Shared connections record high levels of monthly consumption, which can expose them to high prices when marginal prices increase with usage .

%%% OLD

@@

This section describes an extension of the model to allow for exogenous interruptions in a water source that induce households to substitute toward another water source.  Exogenous interruptions produce useful variation in the types of water sources used by each household, serving an important role in estimating the marginal costs of transporting water from a neighbor.  Due to shifts in the geology, road construction, or aging infrastructure, households sometimes experience underground breaks in the service pipe between the main line and the water meter.  These leaks result in repair costs and water bills borne by households that can be large enough to force affected households to switch to an alternative water source.

@@

This paper demonstrates that pricing policies intended to increase water access especially among the poor can have unintended consequences when low-income households can simply access the piped water network through their neighbors.

The counterfactual exercises find that a simple two-part tariff structure outperforms the increasing block tariff currently implemented in Manila in terms of overall access to water, total welfare, and even welfare among the bottom half of water users.  In this counterfactual, high fixed costs of accessing an individual connection allow for sharing networks to play an important role in extending water access despite sizable hassle costs.  I also find that complex non-linear price schedules are largely ineffective in terms of achieving both efficiency and equity goals when sharing networks are readily available.
